---
title: "Boruta Feature Importance Analysis"
author: "Jim Thompson"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup,message=FALSE,warning=FALSE}
library(caret)
library(data.table)
library(Boruta)
library(plyr)
library(dplyr)
library(pROC)

ROOT.DIR <- ".."

ID.VAR <- "Id"
TARGET.VAR <- "SalePrice"

```
This report determines what features may be relevant to predicting house sale price. This analysis is based on the [**Boruta**](https://cran.r-project.org/web/packages/Boruta/index.html) package.

### Data Preparation for Bourta Analysis
```{r DataRetrieval}

# retrive data for analysis
sample.df <- read.csv(file.path(ROOT.DIR,"input/train.csv"),stringsAsFactors = FALSE)
```

Determine data type for each candidate explanatory attributes.
```{r QuickDataProfile}
# extract only candidate feture names
candidate.features <- setdiff(names(sample.df),c(ID.VAR,TARGET.VAR))
data.type <- sapply(candidate.features,function(x){class(sample.df[[x]])})
table(data.type)

print(data.type)

# deterimine data types
explanatory.attributes <- setdiff(names(sample.df),c(ID.VAR,TARGET.VAR))
data.classes <- sapply(explanatory.attributes,function(x){class(sample.df[[x]])})

# categorize data types in the data set?
unique.classes <- unique(data.classes)

attr.data.types <- lapply(unique.classes,function(x){names(data.classes[data.classes==x])})
names(attr.data.types) <- unique.classes

```

Prepare data set for Boruta analysis.  For this analysis, missing values are
handled as follows:
* missing numeric data is set to -1
* missing character data is set to __*MISSING*__

For this analysis character attributes are handled as factors.  One limitation of 
handling character data as factors is the underlying assumption there is a relative 
ordering for the factor levels.
```{r PrepareData}
# pull out the response variable
response <- sample.df$SalePrice

# remove identifier and response variables
sample.df <- sample.df[candidate.features]

# for numeric set missing values to -1 for purposes of the random forest run
for (x in attr.data.types$integer){
  sample.df[[x]][is.na(sample.df[[x]])] <- -1
}

for (x in attr.data.types$character){
  sample.df[[x]][is.na(sample.df[[x]])] <- "*MISSING*"
}

```



### Run Boruta Analysis
```{r borutarun}

set.seed(13)
bor.results <- Boruta(sample.df,response,
                   maxRuns=101,
                   doTrace=0)
```


### Boruta results
```{r borutaresults1,echo=FALSE}
cat("\nSummary of Boruta run:\n")
print(bor.results)
```

These attributes were deemed as relevent to predicting house sale price.
```{r echo=FALSE}
cat("\n\nRelevant Attributes:\n")
getSelectedAttributes(bor.results)
```

The following plot shows the relative importance of each candidate explanatory attribute.
The x-axis represents each of candidate explanatory variables.  Green color indicates
the attributes that are relevant to prediction.  Red indicates attributes that
are not relevant.  Yellow color indicates attributes that may or may not be relevant to 
predicting the response variable.
```{r borutaplot,echo=FALSE,fig.width=9,fig.height=7}
plot(bor.results)
```

Detailed results for each candidate explanatory attributes.
```{r detailattrdump,echo=FALSE}
cat("\n\nAttribute Importance Details:\n")
options(width=125)
arrange(cbind(attr=rownames(attStats(bor.results)), attStats(bor.results)),desc(medianImp))
```
